Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-04-20 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
updates to appendices and fig captions

\begin{figure}[h]
\includegraphics[width=\columnwidth]{figs/edgb_gamma.pdf}
\caption{(Color online).~Posterior distribution function on the dimensionless parameter $\gamma_{\rm EdGB}$,
defined in Eq.~\eqref{eq:def_gamma_gb}. We see that this parameter controls the ParSpec expansion in EdGB gravity
does have maximum support at $\gamma_{\rm EdGB} = 0$. This shows that our model is consistent with GR.}
\label{fig:gamma_gb}
\end{figure}

@@

In this Appendix, we show how we determined the coefficients in the ParSpec
using QNM calculations found in the literature for the theories described in
Sec.~\ref{sec:review_theories}.

@@

We start by considering EdGB gravity, for which the QNMs of nonrotating and
slowly-rotating BHs where calculated in
Refs.~\cite{Pani:2009wy,Blazquez-Salcedo:2016enn} and~\cite{Pierini:2021jxd},
respectively.

@@

% Let us show one concrete example of how this mapping is done.
%
In Ref.~\cite{Wagle:2021tam} the QNMs of slowly-rotating BHs in dCS gravity were numerically determined
and it was found that the axial gravitational modes have their damping time increased as we
increase the Chern-Simons coupling $\ell_{\rm CS}$ at fixed spin $\chi$ of the BH.
%
Hence, according to the prescription of Sec.~\ref{sec:theory_specific_qnm}, this is the branch of QNMs we will select.

We can now proceed to determine $\delta\omega^{(j)}$ and $\delta\tau^{(j)}$ as follows.
%
Using the fitting formula Eq.~(54a) of~\cite{Wagle:2021tam}, namely,
%
\begin{equation}
    M_{\rm f} \omega_{\rm CS} = c_{1} + c_{2} \kappa \zeta + (c_{3} + c_{4} \kappa \zeta) \, (1 - \chi)^{c_{5} + c_{6} \kappa \zeta},
    \label{eq:omega_dsc_eg}
\end{equation}
%
and similarly for the imaginary part ${\rm Im}(\sigma_{\rm CS}) =  - 1 / \tau_{\rm CS}$.
%
Here $\kappa = 1/(16 \pi)$, $\zeta = \ell_{\rm CS}^{4} / (M_{\rm f}^4 \kappa)$, and thus
%
\begin{equation}
    \kappa \zeta = (\ell_{\rm CS} / M_{\rm f})^4 \,,
\end{equation}
%
and $c_{i}$ are fitting coefficients which can be found in Ref.~\cite{Wagle:2021tam}, Table II.

We now expand Eq.~\eqref{eq:omega_dsc_eg} to leading-orders in $\chi$ and $\ell_{\rm CS}$, and
gather the terms proportional to $\ell_{\rm CS}$.
%
We obtain
%
\begin{align}
    M_{\rm f} \, \omega_{\rm CS} &=
    \left[ 0.3722 + 1.1945 (\ell_{\rm CS} / M_{\rm f})^4 \right]
    \nonumber \\
    &\quad + \left[0.1861 + 5.1828 (\ell_{\rm CS} / M_{\rm f})^4 \right] \, \chi\,,
    \label{eq:omega_dcs_interm}
\end{align}
%
where we made use of the numerical value of the coefficients $c_{i}$.
%
We find that (reassuringly) that the nonrotating GR part of the expression above agree with $\omega^{(0)}$ of Ref.~\cite{Maselli:2019mjd}
to 0.5\% relative percent error. The same error estimate is larger ($\approx 20$\%) for the linear-in-spin coefficient;
0.1861 in comparison to 0.1258 of Ref.~\cite{Maselli:2019mjd}.
%
We attribute this difference to Ref.~\cite{Wagle:2021tam} having fitted Eq.~\eqref{eq:omega_dsc_eg} to QNM data compute to linear-order in spin, whereas~\cite{Maselli:2019mjd}
fitted Eq.~\eqref{eq:kerr_expansion} to Kerr QNM valid to all orders in spin.

We can now isolate the dCS corrections from Eq.~\eqref{eq:omega_dcs_interm} and
compare against Eq.~\eqref{eq:qnm_nongr_iso}, to find
%
\begin{equation}
p_{\rm CS} = 4\,, \quad \delta \omega^{(0)}_{\rm CS} = 3.1964\,, \quad \delta \omega^{(1)}_{\rm CS} = 41.199\,.
\label{eq:cs_omega_coefs}
\end{equation}

We can carry the same steps for $\tau_{\rm CS}$ and find
%
\begin{equation}
\delta \tau^{(0)}_{\rm CS} = 6.3619\,, \quad \delta \tau^{(1)}_{\rm CS} = 794.66\,.
\label{eq:cs_tau_coefs}
\end{equation}
%
which completes the set of fixed parameters beyond-GR parameters in the ringdown of the \pSEOB{} waveform model.
%
We remark that the alarmingly large values of $\delta \omega_{\rm CS}^{(1)}$ and $\delta \tau_{\rm CS}^{(1)}$
are compensated by the assumptions that $(\ell_{\rm CS} / M_{\rm f})^2$ and $\chi$ are much less than unity. These
are the assumptions used to calculate the QNM frequencies in Ref.~\cite{Wagle:2021tam}.

>>>>

\begin{figure}[h]
\includegraphics[width=\columnwidth]{figs/edgb_gamma.pdf}
\caption{(Color online).~Posterior distribution function on the dimensionless parameter $\gamma_{\rm EdGB}$,
defined in Eq.~\eqref{eq:def_gamma_gb}. Different line colors distinguish between events, while
different line styles distinguish between different $N_{\rm max}$.
%
We see that this parameter which controls the ParSpec expansion in EdGB gravity
does have maximum support at $\gamma_{\rm EdGB} = 0$. This shows that our model is consistent with GR.
\hs{The right panel shows the PDFs in the range $0 \leqslant \gamma_{\rm EdGB} \leqslant 1$. Note how
the curves are flat in this range for both events and when $N_{\rm max} = 1$.}
}
\label{fig:gamma_gb}
\end{figure}

@@

In this Appendix, we show how we determined the coefficients in the ParSpec
and summarized in Table~\ref{tab:ref_theories_qnms},
using QNM calculations found in the literature for the theories described in
Sec.~\ref{sec:review_theories}.
%
Here, we consider only the fundamental QNM $(\ell, m, n) = (2, 2, 0)$ and
hence we omit the QNM subscript ``$(2,2,0)$'' for brevity.

@@

We start by considering EdGB gravity, for which the QNMs of nonrotating and
slowly-rotating BHs where calculated in Refs.~\cite{Pani:2009wy,Blazquez-Salcedo:2016enn,Blazquez-Salcedo:2017txk}
and~\cite{Pierini:2021jxd}, respectively.
%
In particular, Ref.~\cite{Blazquez-Salcedo:2016enn} found that the damping time
of the polar gravitational-led modes increases as the coupling constant
$\ell_{\rm EdGB}$ is increased.
%
Hence, according to the prescription of Sec.~\ref{sec:theory_specific_qnm},
this is the gravitational-led branch of QNMs we select.

For the nonrotating QNMs we used the numerical data of Ref.~\cite{Blazquez-Salcedo:2016enn} and
generated a new quartic in $\ell_{\rm EdCS}$ fit to the data valid for small
values of the coupling constant [see, in particular, Eq.~(27) and Fig.~2 of~\cite{Blazquez-Salcedo:2016enn}].
%
We find,
%
\begin{subequations}
\begin{align}
    M {\rm Re}(\sigma)_{\rm EdGB} &= M {\rm Re}(\sigma)_{\rm GR} \, \left[ 1 + 0.0107 (\ell_{\rm EdGB} / M)^4 \right],
    \\
    M {\rm Im}(\sigma)_{\rm EdGB} &= M {\rm Im}(\sigma)_{\rm GR} \, \left[ 1 - 0.0044 (\ell_{\rm EdGB} / M)^4 \right].
\end{align}
\end{subequations}
%
The smallness of the numerical prefactors of $\ell_{\rm EdGB}^4$ are a consequence of the how weakly the QNMs
of BHs in EdGB gravity deviate from their GR counterparts, even at moderate values $(\ell_{\rm EdGB} / M)^4 \approx 0.3$.

The rotating corrections are included using the results of Ref.~\cite{Pierini:2021jxd}, in particular their Eqs.~(51) and~(52),
which for consistency, with our previous discussion, we truncated at quartic-order in $\ell_{\rm EdGB}$.
%
Taken together, all use as our fitting expressions for the QNM is EdGB gravity
%
\begin{subequations}
\begin{align}
    M {\rm Re}(\sigma)_{\rm EdGB} &= \dots % {\rm Re}(\omega)_{\rm GR} \, \left[ 1 + 0.0107 (\ell_{\rm EdGB} / M)^4 \right],
    \\
    M {\rm Im}(\sigma)_{\rm EdGB} &= \dots % {\rm Im}(\omega)_{\rm GR} \, \left[ 1 - 0.0044 (\ell_{\rm EdGB} / M)^4 \right].
\end{align}
\end{subequations}

\dots

@@

For dCS gravity, we follow Ref.~\cite{Wagle:2021tam} which numerically calculated the QNMs of slowly-rotating BHs in this theory
and found that the axial gravitational modes have their damping time increased as we
increase the Chern-Simons coupling $\ell_{\rm dCS}$ at fixed spin $\chi$ of the BH.
%
Hence, according to the prescription of Sec.~\ref{sec:theory_specific_qnm},
this is the branch of QNMs we will select.

We can now proceed to determine $\delta\omega^{(j)}$ and $\delta\tau^{(j)}$ as follows.
%
Using the fitting formula Eq.~(54a) of~\cite{Wagle:2021tam}, namely,
%
\begin{equation}
    M \omega_{\rm CS} = c_{1} + c_{2} \kappa \zeta + (c_{3} + c_{4} \kappa \zeta) \, (1 - \chi)^{c_{5} + c_{6} \kappa \zeta},
    \label{eq:omega_dsc_eg}
\end{equation}
%
and similarly for the imaginary part ${\rm Im}(\sigma_{\rm dCS}) =  - 1 / \tau_{\rm dCS}$.
%
Here $\kappa = 1/(16 \pi)$, $\zeta = \ell_{\rm dCS}^{4} / (M^4 \kappa)$, and thus
%
\begin{equation}
    \kappa \zeta = (\ell_{\rm dCS} / M)^4 \,,
\end{equation}
%
and $c_{i}$ are fitting coefficients which can be found in Ref.~\cite{Wagle:2021tam}, Table II.

We now expand Eq.~\eqref{eq:omega_dsc_eg} to leading-orders in $\chi$ and $\ell_{\rm dCS}$, and
gather the terms proportional to $\ell_{\rm dCS}$.
%
We obtain
%
\begin{align}
    M \, \omega_{\rm dCS} &=
    \left[ 0.3722 + 1.1945 (\ell_{\rm dCS} / M)^4 \right]
    \nonumber \\
    &\quad + \left[0.1861 + 5.1828 (\ell_{\rm dCS} / M)^4 \right] \, \chi\,,
    \label{eq:omega_dcs_interm}
\end{align}
%
where we made use of the numerical value of the coefficients $c_{i}$.
%
We find that (reassuringly) that the nonrotating GR part of the expression above agree with $\omega^{(0)}$ of Ref.~\cite{Maselli:2019mjd}
to 0.5\% relative percent error. The same error estimate is larger ($\approx 20$\%) for the linear-in-spin coefficient;
0.1861 in comparison to 0.1258 of Ref.~\cite{Maselli:2019mjd}.
%
We attribute this difference to Ref.~\cite{Wagle:2021tam} having fitted Eq.~\eqref{eq:omega_dsc_eg} to QNM data compute to linear-order in spin, whereas~\cite{Maselli:2019mjd}
fitted Eq.~\eqref{eq:kerr_expansion} to Kerr QNM valid to all orders in spin.

We can now isolate the dCS corrections from Eq.~\eqref{eq:omega_dcs_interm} and
compare against Eq.~\eqref{eq:qnm_nongr_iso}, to find
%
\begin{equation}
p_{\rm dCS} = 4\,, \quad \delta \omega^{(0)}_{\rm dCS} = 3.1964\,, \quad \delta \omega^{(1)}_{\rm dCS} = 41.199\,.
\label{eq:cs_omega_coefs}
\end{equation}

We can carry the same steps for $\tau_{\rm dCS}$ and find
%
\begin{equation}
\delta \tau^{(0)}_{\rm dCS} = 6.3619\,, \quad \delta \tau^{(1)}_{\rm dCS} = 794.66\,.
\label{eq:cs_tau_coefs}
\end{equation}
%
which completes the set of fixed parameters beyond-GR parameters in the ringdown of the \pSEOB{} waveform model.
%
We remark that the alarmingly large values of $\delta \omega_{\rm dCS}^{(1)}$ and $\delta \tau_{\rm dCS}^{(1)}$
are compensated by the assumptions that $(\ell_{\rm dCS} / M_{\rm f})^2$ and $\chi$ are much less than unity,
which are the assumptions in Ref.~\cite{Wagle:2021tam}.
